---
title: "SwissRegulon and UniPROBE"
author: "Matt Richards"
date: "7/18/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Motivation

Up to this point, we've looked exclusively at creating our set of motifs from Jaspar and HOCOMOCO; however, our investigation has revealed that using only these motifs results in missing 78 transcription factors (TFs) in Seth's original set. It isn't all bad; we're also adding something like 400 TFs that he doesn't have. Still, we think it's vital to make sure we cover all his TFs, but at the same time we don't want to arbitrarily just add motifs. So what do we do?

Well, to the ends of figuring out "what to do", we've performed TOMTOM again, this time on the SwissRegulon motifs and the UniPROBE motifs as query sequences, with the existing set of Jaspar/HOCOMOCO motifs as targets. In this document, we'll explore the implications of these TOMTOM runs, specifically looking to answer the following questions:

1. How many motifs would we be adding from each database, disregarding the fact that some of these motifs will overlap with one another?
2. How many TFs would we be adding from each database? 
3. What is the TF overlap with both each other and with the existing Jaspar/HOCOMOCO motifs?

We will use this markdown document to address these questions and possibly more as they come up. 

## Grabbing and Formatting the Results

As with our previous mapping exercise, we'll pull out the TOMTOM results and stick them in data frames:

```{r}
uniprobe <- read.table(file = "./uniprot_compare/tomtom.txt")
swissreg <- read.table(file = "./swiss_regulon_compare/tomtom.txt")
names(uniprobe) <- c("Query.ID", "Target.ID", "Optimal.offset", "Pvalue","Evalue","Qvalue","Overlap","Query.consensus","Target.consensus","Orientation")
names(swissreg) <- c("Query.ID", "Target.ID", "Optimal.offset", "Pvalue","Evalue","Qvalue","Overlap","Query.consensus","Target.consensus","Orientation")
```

Here's some stats on what we just pulled out:
```{r}
dim(uniprobe); dim(swissreg)
```

Notably, there's many more (about 2-2.5x as many) hits than before. This is a little surprising for UniPROBE, which only had 380 motifs to use as query sequences, but then again we're matching against many more motifs than before. So this is probably fine. 

Starting with UniPROBE and then doing the same for SwissRegulon, let's screen out only those values below our P-value threshold from the previous analysis:

```{r}
above.uniprobe <- uniprobe %>% 
  select(Query.ID, Target.ID, Pvalue) %>%
  mutate(logP = -1*log10(Pvalue)) %>% filter(logP >= 7.3)

above.uniprobe$Query.ID <- as.character(above.uniprobe$Query.ID)
above.uniprobe$Target.ID <- as.character(above.uniprobe$Target.ID)
repeat.uniprobe <- unique(above.uniprobe$Query.ID)
length(repeat.uniprobe)
```

Looks like of the 380 motifs in UniPROBE, we'd find hits for 218, meaning we'd add 162 to our existing set. Now for SwissRegulon, which has 684 total motifs:

```{r}
above.swiss <- swissreg %>% 
  select(Query.ID, Target.ID, Pvalue) %>%
  mutate(logP = -1*log10(Pvalue)) %>% filter(logP >= 7.3)

above.swiss$Query.ID <- as.character(above.swiss$Query.ID)
above.swiss$Target.ID <- as.character(above.swiss$Target.ID)
repeat.swiss <- unique(above.swiss$Query.ID)
length(repeat.swiss)
```

The news is perhaps even more encouraging for SwissRegulon, which would find repeats for 558 of 684, meaning we'd only add 126 motifs to the set. This doesn't even consider that some of the SwissRegulon might match up with UniPROBE as well, so rather than 288, we might add fewer. The 288 possible motifs really represent the upper limit. 

***RESERVED FOR FUTURE SWISSREG/UNIPROBE COMPARISON***

We can investigate how close we are to that limit as well by loading a TOMTOM of SwissRegulon as the query and UniPROBE as the target. We'll have to compare these to the `repeat.swiss` motifs to see how many are the same as UniPROBE.

```{r}
swiss.v.uniprobe <- read.table(file = "./swiss_vs_uniprobe/tomtom.txt")
names(swiss.v.uniprobe) <- c("Query.ID", "Target.ID", "Optimal.offset", "Pvalue","Evalue","Qvalue","Overlap","Query.consensus","Target.consensus","Orientation")

above.swiss.v.uniprobe <- swiss.v.uniprobe %>% 
  select(Query.ID, Target.ID, Pvalue) %>%
  mutate(logP = -1*log10(Pvalue)) %>% filter(logP >= 7.3)

above.swiss.v.uniprobe$Query.ID <- as.character(above.swiss.v.uniprobe$Query.ID)
above.swiss.v.uniprobe$Target.ID <- as.character(above.swiss.v.uniprobe$Target.ID)
repeat.swiss.v.uniprobe <- unique(above.swiss.v.uniprobe$Query.ID)
length(repeat.swiss.v.uniprobe)
```

Of the 684 motifs in SwissRegulon, it looks like we could actually represent 215 with UniPROBE. To get our final count, we need to cross-reference these with the 126 motifs we'd be adding to Jaspar/HOCOMOCO. So let's go ahead and grab the lists of 162 and 126 motifs we'd be adding from UniPROBE and SwissRegulon, respectively. As before, we'll likely have to do some ID trimming

```{r message = FALSE}
library(MotifDb)
all.uniprobe <- query(MotifDb, "uniprobe")
all.swiss <- query(MotifDb, "swissregulon")
length(all.uniprobe); length(all.swiss)
```
Looks like the correct length; let's see what the IDs look like and how they compare to what we're looking for, using UniPROBE first:

```{r}
head(values(all.uniprobe)$providerId)
head(repeat.uniprobe)
```

For the uniprobe, we'll have to pull out only the "UP####"

```{r}
trimmed.uniprobe <- gsub(".*(UP\\d{5}).*","\\1",repeat.uniprobe)
uniprobe.to.add <- subset(all.uniprobe, !(providerId %in% trimmed.uniprobe))
length(uniprobe.to.add)
```

Excellent; now let's do the same with SwissRegulon:

```{r}
head(values(all.swiss)$providerId)
head(repeat.swiss)
```

Again, we're a bit off. We'll trim off what we want here as well:

```{r}
trimmed.swiss <- gsub("NA-SwissRegulon-(.+)","\\1",repeat.swiss)
swiss.to.add <- subset(all.swiss, !(providerId %in% trimmed.swiss))
length(swiss.to.add)
```

Beautiful! Let's do the same for the IDs that matched UniPROBE and see how many of these overlap:

```{r}
trimmed.overlaps <- gsub("NA-SwissRegulon-(.+)","\\1",repeat.swiss.v.uniprobe)
swiss.in.uniprobe <- intersect(trimmed.overlaps, values(swiss.to.add)$providerId)
length(swiss.in.uniprobe)
```

Alright, so let's tie a bow on this whole thing and summarize the motifs:

1. UniPROBE would add 162 motifs that are distinct from those already in our FIMO
2. SwissRegulon would add 126 motifs that are distinct from those already in our FIMO
3. 38 of the 126 these SwissRegulon motifs are basically the same as some from UniPROBE

Based upon this summary, we are pointed toward the following:

1. Add the 162 UniPROBE motifs and 88 unique SwissRegulon ones, totaling **250** new motifs
2. Add mappings between Jaspar/HOCOMOCO and non-unique motifs they matched to
3. Add mappings between the 38 motifs that matched between UniProbe and SwissRegulon

Toward this end, let's section off the SwissRegulon motifs we actually want to add:

```{r}
final.swiss.to.add <- subset(swiss.to.add, !(providerId %in% swiss.in.uniprobe))
length(final.swiss.to.add)
```

We now have 2 motif files to add: `uniprobe.to.add` and `final.swiss.to.add`. We should also take the time now to construct our mappings now; these are steps 2 and 3 above. We'll add these 2 sets together as well:

```{r}
all.new.motifs <- c(uniprobe.to.add, final.swiss.to.add)
```


## Adding missing Jaspar Motifs

There's one more source of motifs we haven't yet added: the missing Jaspar motifs for other vertebrates. Using the code we've already written, we can pull these out as well:

```{r}
vertebrates <- c("Ggallus","Nsp.","Ocuniculus","Rnorvegicus","Rrattus","Vertebrata", "Xlaevis")
new.motifs <- subset(query(MotifDb, "jaspar2016"), organism %in% vertebrates)
length(new.motifs)
```

We'll finish by adding these to our final list of motifs to add; this should finish it off and give us 274 total motifs to add:

```{r}
final.new.motifs <- c(all.new.motifs, new.motifs)
length(final.new.motifs)
```

And finally, let's save these motifs as both an RDS and a .MEME file for later use

```{r}
export(final.new.motifs, "../2017_07_26_new_motifs.meme",format = "meme")
saveRDS(final.new.motifs, "../2017_07_26_new_motifs.RDS")
```

## Mapping the Final Motifs

As in our previous exercise, we have 2 types of assets we're going to put together here: mappings and our final list of motifs. However, there are several mappings so here is what we actually have

1. `above.uniprobe`: The TOMTOM mappings of UniPROBE to Jaspar/HOCOMOCO
2. `above.swiss`: The TOMTOM mappings of SwissRegulon to Jaspar/HOCOMOCO
3. `above.swiss.v.uniprobe` The TOMTOM mappings of SwissReguon to UniProbe
4. `

## Extracting TF information

Just as we did before, we need to figure out the TFs associated with these motifs. We'll also want to reload the motifs in our previous mapping and the motifs in Seth's initial mapping. First, we'll do the process newly with our

As for the Jaspar/HOCOMOCO motifs, we saved these previously so we can just load them up:

```{r}
existing.motifs <- readRDS("../jaspar_hocomoco_mapping.RDS")
```

We can also load up Seth's table:
```{r}
seth.table <- read.table(system.file(package="trena", "extdata", "motifGenes.tsv"), sep = "\t", header=TRUE)
```

Annoyingly, these tables have similar info but are somewhat disparate in their column names. We'll quickly fix that so they line up better:

```{r}
names(existing.motifs) <- c("motif","tf.gene")
```

Finally, let's get some perspective on all the things we want to add by finding all the missing motifs and TFs:

```{r}
missing.motifs <- setdiff(seth.table$motif,existing.motifs$motif)
missing.tfs <- setdiff(seth.table$tf.gene,existing.motifs$tf.gene)
```



## How many motifs are we adding?

## How many TFs are we adding? 
